---
title: "6: Part 1 - Generalized Linear Models"
author: "Environmental Data Analytics | John Fay and Luana Lima | Developed by Kateri Salk"
date: "Spring 2022"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Objectives
1. Answer questions on M5/A5
2. Answer questions on M6 - GLMs
3. Additional comments on videos - t-test
4. Practice more application GLM to real datasets


## Set up
```{r, message = FALSE}
library(tidyverse)
library(agricolae)   

PeterPaul.chem.nutrients <- read.csv("./Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv", stringsAsFactors = TRUE)
# Set date to date format
PeterPaul.chem.nutrients$sampledate <- as.Date(PeterPaul.chem.nutrients$sampledate, format = "%Y-%m-%d")

EPAair <- read.csv("./Data/Processed/EPAair_O3_PM25_NC1819_Processed.csv", stringsAsFactors = TRUE)
# Set date to date format
EPAair$Date <- as.Date(EPAair$Date, format = "%Y-%m-%d")


Litter <- read.csv("./Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv", stringsAsFactors = TRUE)
# Set date to date format
Litter$collectDate <- as.Date(Litter$collectDate , format = "%Y-%m-%d")

# Set theme
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)
```

## T-Test

Continuous response, one categorical explanatory variable with two categories (or comparison to a single value if a one-sample test).

### Formulating Hypothesis for µ

Two hypotheses are formed – the null hypothesis and the alternative hypothesis.
The null hypothesis and the alternative hypothesis combine to cover all possible values for the population mean.
The null hypothesis must have the equality.
The null and alternative hypotheses are always stated in terms of the population mean (mu).

### One-sample t-test
The object of a one sample test is to test the null hypothesis that the mean of the group is equal to a specific value. For example, we might ask ourselves (from the EPA air quality processed dataset): 

Function t.test()
**x**	a (non-empty) numeric vector of data values.
**alternative** a character string specifying the alternative hypothesis, must be one of "two.sided" (default), "greater" or "less". You can specify just the initial letter.
**mu** a number indicating the true value of the mean (or difference in means if you are performing a two sample test).
**formula** a formula of the form lhs ~ rhs where lhs is a numeric variable giving the data values and rhs either 1 for a one-sample or paired test or a factor with two levels giving the corresponding groups. If lhs is of class "Pair" and rhs is 1, a paired test is done.


Are Ozone levels below the threshold for "good" AQI index (0-50)?


> Exercise 1: State the hypotheses for testing mean of AQI index.

> Answer: Null is ozone levels below threshold and alternate is levels re above.


```{r}


summary(EPAair$Ozone)
EPAair.subsample <- sample_n(EPAair, 5000)

# Evaluate assumption of normal distribution, shapiro test tells us not normal
shapiro.test((EPAair.subsample$Ozone))
ggplot(EPAair, aes(x = Ozone)) +
  geom_histogram() 

#can see some points way off the line probs why shapiro said not normal
qqnorm(EPAair$Ozone); qqline(EPAair$Ozone)

#did the opposite than what we listed above. null is above here
#significant so show mu is below the threshold so air is good!
O3.onesample <- t.test(EPAair$Ozone, mu = 50, alternative = "less")
O3.onesample

Ozone.plot <- ggplot(EPAair, aes(x = Ozone)) +
  #geom_density(stat = "count", fill = "gray") +
  geom_density(fill = "gray") +
  geom_vline(xintercept = 50, color = "#238b45", lty = 2, size = 0.9) +
  scale_x_continuous(expand = c(0, 0)) + scale_y_continuous(expand = c(0, 0))
print(Ozone.plot)
```

Write a sentence or two about the results of this test. Include both the results of the test and an interpretation that puts the findings in context of the research question.

> use hypothesis test to support a claim about a sample.

### Two-sample t-test

The two-sample *t* test is used to test the hypothesis that the mean of two samples is equivalent. Unlike the one-sample tests, a two-sample test requires a second assumption that the variance of the two groups is equivalent. Are Ozone levels different between 2018 and 2019?

Similar to analysis of variance (ANOVA)

```{r}
shapiro.test(EPAair$Ozone[EPAair$Year == 2018])
shapiro.test(EPAair$Ozone[EPAair$Year == 2019])

#p-value less than 0.05 then reject null for 2018 and 2019 i.e. data do not follow normal distribution

#Compare variance using F-test (only)
var.test(EPAair$Ozone ~ EPAair$Year)

#p-value less than 0.05 then reject null for 2018 and 2019 i.e. true ratio not equal to one

# Format as a t-test, categorical variable with two levels (2018 and 2019)
#reject null b/c p-value below 0.05, means are not the same
O3.twosample <- t.test(EPAair$Ozone ~ EPAair$Year)
O3.twosample
O3.twosample$p.value

# Format as a GLM
#get same p-value as with t-test
O3.twosample2 <- lm(EPAair$Ozone ~ EPAair$Year)
summary(O3.twosample2)

#can see very separate 
plot(O3.twosample2)

#can see very close except for second peak
ggplot(EPAair, aes(x = Ozone, color = as.factor(Year))) +
  geom_freqpoly()


```


## Statistical Test: Cheat sheet

**F-test:** Compare the variances of two groups. The data must be normally distributed.

**Bartlett’s test:** Compare the variances of two or more groups. The data must be normally distributed.

**Shapiro.test:** check for normality

**One-sample t-test:** check if mean is equal/less/greater to specific value, single variable

**Two-sample t-test:**  check if mean of two samples is equivalent

### Visualization and interpretation challenge

Create three plots, each with appropriately formatted axes and legends. Choose a non-default color palette.

1. geom_density of ozone divided by year (distinguish between years by adding transparency to the geom_density layer).
2. geom_boxplot of ozone divided by year . Add letters representing a significant difference between 2018 and 2019 (hint: stat_summary). 
3. geom_violin of ozone divided by year, with the 0.5 quantile marked as a horizontal line. Add letters representing a significant difference between 2018 and 2019. 

```{r}
#Exercise 2:

#1_density
#fill instead of color b/c want whole thing filled with color not jus tthe outline, also want to be able to use tranparency
#have the legend fill = "" blnak in the quotes so the legen doesn't have a title
#as.factor changes year from integer to character, so you can plot it as a categorical variable. only changes it to a character for the ggplot not in the dataset

ozone_year_density <- ggplot(EPAair, aes(x = Ozone, fill = as.factor(Year))) +
  geom_density(aes(color = "gray", alpha = 50))

#2_boxplot

boxplot <- ggplot()

stat_summary(geom)

#the stat_summary function geom means you want text, fun.y means you want the label at the max y value. fun.y needs to be a predetermined function in R either preprogrammed or made by you above this code block, label is where you write what label you want 
#vjust means vertical adjustment

#3_violin


```

## Linear Regression

Important components of the linear regression are the correlation and the R-squared value. The **correlation** is a number between -1 and 1, describing the relationship between the variables. Correlations close to -1 represent strong negative correlations, correlations close to zero represent weak correlations, and correlations close to 1 represent strong positive correlations. The **R-squared value** is the correlation squared, becoming a number between 0 and 1. The R-squared value describes the percent of variance accounted for by the explanatory variables. 

For the NTL-LTER dataset, can we predict PM2.5 from Ozone?

```{r}

#Exercise 3: Run a linear regression PM2.5 by Ozone. Find the p-value and R-squared value. 
Pm2.5byOzone.regression <- lm(EPAair$PM2.5 ~ EPAair$Ozone)

#could also do

#just printing it only returns the coefficients
Pm2.5byOzone.regression

#summary returns more details including p-value, r-squared is very low indicating other things than ozone influencing PM2.5
summary(Pm2.5byOzone.regression)

#Exercise 4: Build a scatterplot. Add a line and standard error for the linear regression. 
scatter.regression <- ggplot(Pm2.5byOzone.regression, aes(x = Ozone, y = PM2.5)) +
  geom_point()
print(scatter.regression)


```


## AIC to select variables

What other variables can we add to improve model?

```{r}

#Exercise 5: Build correlation plots and identify more possible explanatory variables to add to the regression.




#Exercise 6: Choose a model by AIC in a Stepwise Algorithm. Do the results from AIC match the variables you selected on Exercise 5?




#Exercise 7: Run another regression using the variables selected on Exercise 6. Compare r-squared value with the one from Exercise 5. 

```






